---
title: "Initial Analysis"
author: "Ben Lorentz"
date: "2023-09-30"
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

1. Descriptive statistics

a. Read in Data

```{r load data}
source(".installpackages.R")

library(tidyverse)
library(amt)
library(sf)
library(lubridate) 
library(ggplot2)
library(terra)
library(png)

vulture_metadata <- read.csv("../data/Black-Vultures-and-Turkey-Vultures-Southeastern-USA-reference-data.csv")
print(vulture_metadata)

vulture_dat <- read.csv("../data/Black-Vultures-and-Turkey-Vultures-Southeastern-USA.csv")
head(vulture_dat)
str(vulture_dat)
nrow(vulture_dat)
#vulture_dat
```

b. how many individuals examined

```{r how many individuals in dataset} 
# tag id's 17 tags 

sort(unique(vulture_dat$tag.local.identifier))
length(unique(vulture_dat$tag.local.identifier))

# individual bird id's 18 birds

sort(unique(vulture_dat$individual.local.identifier))
length(unique(vulture_dat$individual.local.identifier))
```

c. How many datapoints are present for each individual in the timeperiod

```{r how many datapoints are present over what time}

bird_ids <- unique(vulture_dat$individual.local.identifier)

record_table <- data.frame()

for(i in 1:length(bird_ids)){
  current_bird <- bird_ids[i]
  number_of_records <- sum(vulture_dat$individual.local.identifier == current_bird)
  new_row <- c(current_bird,number_of_records)
  #colnames(new_row) <- c("bird id","number of records")
  record_table <- rbind(record_table, new_row)
}

colnames(record_table) <-  c("bird id","number of records")

(record_table)
```
d. When does each record start and stop?

```{r find start and stop points for each record}

bird_ids <- unique(vulture_dat$individual.local.identifier)

start_stop_table <- data.frame()

for(i in 1:length(bird_ids)){
  current_bird <- bird_ids[i]
  curr_records <- vulture_dat[vulture_dat$individual.local.identifier == current_bird,]
  start_rec <- head(curr_records, n=1)$timestamp
  stop_rec <- tail(curr_records, n=1)$timestamp
  n_days <- round(as.numeric(difftime(ymd_hms(stop_rec), ymd_hms(start_rec),units = "days")),3)
  new_row <- c(current_bird,start_rec, stop_rec, n_days)
  
  start_stop_table <- rbind(start_stop_table, new_row)
}

colnames(start_stop_table) <-  c("bird id","start","stop","n days")

(start_stop_table)

```


e. Simple plot of the data points for one individual 

```{r plot the data for 91}
#select bird 91

id <- 91

#subset whole dataset

vulture_91 <- vulture_dat[vulture_dat$individual.local.identifier == 91,]

plot(vulture_91$location.long, vulture_91$location.lat)

ggplot(vulture_91, aes(x=location.long, y=location.lat))+ geom_point()

```

2. Estimate home range for one individual using three methods of your own choice.

0. Choose Individual and generate Tracks

```{r select bird id 91}

vulture_91 <- vulture_dat %>%
  filter(individual.local.identifier == 91)

head(vulture_91)

# check timestamp
class(vulture_91$timestamp)

# convert timestamp to posixct format
vulture_91$timestamp <- ymd_hms(vulture_91$timestamp, tz = "UTC")
head(vulture_91$timestamp)
str(vulture_91$timestamp)

# make track for vulture 91
trk_91 <- make_track(vulture_91, location.long,location.lat, timestamp, id = individual.local.identifier, crs = 4326)

# save this file to disk
saveRDS(trk_91, file = "../output/vulture_91_gps_data_track.rds")

# check sampling rate for bird 91
summarize_sampling_rate(trk_91)

# This suggests that the median sampling rate is 2h, but varying up to
# 12h. We can now resample the whole track to 2h interval (with tolerance of
# 10 min), so that if there are more than 2h between relocations, they will 
# be divided into different bursts.

trk_91_resamp <- track_resample(trk_91, rate = minutes(1), tolerance = minutes(1490))

# add step length as a new col
trk_91_sl <- trk_91_resamp %>% mutate(sl = step_lengths(.)) 

# calculate steps by burst
trk_91_sbb <- trk_91_resamp %>% steps_by_burst()
```

a. MCP Home Range

```{r mcp home range estimation for 91}
mcps <- hr_mcp(trk_91, levels = c(0.5, 0.75, 0.95, 1))
mcps

# Get area in km^2
hr_area(mcps, units = TRUE) %>% 
  mutate(area = units::set_units(area, "km^2"))

# plot the polygon isopleths by themeself
plot(hr_isopleths(mcps)[1])

# Plot the isopleth with data on top
plot(mcps) 

# Custom plot with ggplot2 (thanks to Brian Smith for this code chunk)

hr_isopleths(mcps) %>% 
  # Make level a factor for discrete color scales
  # Can control order and labels here
  mutate(level = factor(level, 
                        levels = c("1", "0.95", "0.75", "0.5"),
                        labels = c("100%", "95%", "75%", "50%"))) %>% 
  ggplot() +
  geom_sf(aes(color = level), 
          fill = NA, linewidth = 2) +
  geom_point(data = mcps$data, aes(x = x_, y = y_),
             color = "gray30", alpha = 0.5) +
  xlab(NULL) +
  ylab(NULL) +
  scale_color_brewer(name = "MCP Level",
                     palette = "Set1") +
  theme_bw() +
  theme(legend.position = "bottom", 
        legend.box.background = element_rect(colour = "black", linewidth = 0.8))
```


b. Kernel density estimation of Home Range

```{r KDE home range estimation for 91}

kdes <- hr_kde(trk_91, levels = c(0.5, 0.95),)

# Get area in km^2
hr_area(kdes, units = TRUE) %>% 
  mutate(area = units::set_units(area, "km^2"))

# plot the polygon isopleths by themeself
plot(hr_isopleths(kdes)[1])

# Plot the isopleth with data on top
plot(kdes) 

# Custom plot with ggplot2 (thanks to Brian Smith for this code chunk)

hr_isopleths(kdes) %>% 
  # Make level a factor for discrete color scales
  # Can control order and labels here
  mutate(level = factor(level, 
                        levels = c("1", "0.95", "0.75", "0.5"),
                        labels = c("100%", "95%", "75%", "50%"))) %>% 
  ggplot() +
  geom_sf(aes(color = level), 
          fill = NA, linewidth = 2) +
  geom_point(data = mcps$data, aes(x = x_, y = y_),
             color = "gray30", alpha = 0.5) +
  xlab(NULL) +
  ylab(NULL) +
  scale_color_brewer(name = "MCP Level",
                     palette = "Set1") +
  theme_bw() +
  theme(legend.position = "bottom", 
        legend.box.background = element_rect(colour = "black", linewidth = 0.8))

hr_overlap(mcps,kdes)
```

c. Autocorrelated Kernel Density estimator of home range

```{r}

# Without going into detail on the different CTMMs, we'll demonstrate fitting
# an aKDE with an Ornstein-Uhlenbeck model.
akdes <- hr_akde(trk_91, model = fit_ctmm(trk_91, "iid"), levels = 0.95)

# Get area in km^2
hr_area(akdes, units = TRUE) %>% 
  mutate(area = units::set_units(area, "km^2"))

# plot the polygon isopleths by themeself
plot(hr_isopleths(akdes)[1])

# Plot the isopleth with data on top
plot(akdes) 

# Custom plot with ggplot2 (thanks to Brian Smith for this code chunk)

hr_isopleths(akdes) %>% 
  # Make level a factor for discrete color scales
  # Can control order and labels here
  mutate(level = factor(level, 
                        levels = c("1", "0.95", "0.75", "0.5"),
                        labels = c("100%", "95%", "75%", "50%"))) %>% 
  ggplot() +
  geom_sf(aes(color = level), 
          fill = NA, linewidth = 2) +
  geom_point(data = mcps$data, aes(x = x_, y = y_),
             color = "gray30", alpha = 0.5) +
  xlab(NULL) +
  ylab(NULL) +
  scale_color_brewer(name = "MCP Level",
                     palette = "Set1") +
  theme_bw() +
  theme(legend.position = "bottom", 
        legend.box.background = element_rect(colour = "black", linewidth = 0.8))

hr_overlap(kdes,akdes)

```

### Examine Bird 108


e. Simple plot of the data points for one individual 

```{r plot the data}
#select bird 108

id <- 108

#subset whole dataset

vulture_108 <- vulture_dat[vulture_dat$individual.local.identifier == 108,]

plot(vulture_108$location.long, vulture_108$location.lat)

ggplot(vulture_108, aes(x=location.long, y=location.lat))+ geom_point()

```


2. Estimate home range for one individual using three methods of your own choice.

0. Choose Individual and generate Tracks

```{r select bird id 108}

vulture_108 <- vulture_dat %>%
  filter(individual.local.identifier == 108)

head(vulture_108)

# check timestamp
class(vulture_108$timestamp)

# convert timestamp to posixct format
vulture_108$timestamp <- ymd_hms(vulture_108$timestamp, tz = "UTC")
head(vulture_108$timestamp)
str(vulture_108$timestamp)

# make track for vulture 91
trk_108 <- make_track(vulture_108, location.long,location.lat, timestamp, id = individual.local.identifier, crs = 4326)

# save this file to disk
saveRDS(trk_108, file = "../output/vulture_108_gps_data_track.rds")

# check sampling rate for bird 108
summarize_sampling_rate(trk_108)

# This suggests that the median sampling rate is 2h, but varying up to
# 12h. We can now resample the whole track to 2h interval (with tolerance of
# 10 min), so that if there are more than 2h between relocations, they will 
# be divided into different bursts.

trk_108_resamp <- track_resample(trk_108, rate = minutes(2), tolerance = minutes(300))

# add step length as a new col
trk_108_sl <- trk_108_resamp %>% mutate(sl = step_lengths(.)) 

# calculate steps by burst
trk_108_sbb <- trk_108_resamp %>% steps_by_burst()
```

a. MCP Home Range

```{r mcp home range estimation}
mcps <- hr_mcp(trk_108, levels = c(0.5, 0.75, 0.95, 1))
mcps

# Get area in km^2
hr_area(mcps, units = TRUE) %>% 
  mutate(area = units::set_units(area, "km^2"))

# plot the polygon isopleths by themeself
plot(hr_isopleths(mcps)[1])

# Plot the isopleth with data on top
plot(mcps) 

# Custom plot with ggplot2 (thanks to Brian Smith for this code chunk)

hr_isopleths(mcps) %>% 
  # Make level a factor for discrete color scales
  # Can control order and labels here
  mutate(level = factor(level, 
                        levels = c("1", "0.95", "0.75", "0.5"),
                        labels = c("100%", "95%", "75%", "50%"))) %>% 
  ggplot() +
  geom_sf(aes(color = level), 
          fill = NA, linewidth = 2) +
  geom_point(data = mcps$data, aes(x = x_, y = y_),
             color = "gray30", alpha = 0.5) +
  xlab(NULL) +
  ylab(NULL) +
  scale_color_brewer(name = "MCP Level",
                     palette = "Set1") +
  theme_bw() +
  theme(legend.position = "bottom", 
        legend.box.background = element_rect(colour = "black", linewidth = 0.8))
```

b. Kernel density estimation of Home Range

```{r KDE home range estimation}

kdes <- hr_kde(trk_108, levels = c(0.5, 0.95),)

# Get area in km^2
hr_area(kdes, units = TRUE) %>% 
  mutate(area = units::set_units(area, "km^2"))

# plot the polygon isopleths by themeself
plot(hr_isopleths(kdes)[1])

# Plot the isopleth with data on top
plot(kdes) 

# Custom plot with ggplot2 (thanks to Brian Smith for this code chunk)

hr_isopleths(kdes) %>% 
  # Make level a factor for discrete color scales
  # Can control order and labels here
  mutate(level = factor(level, 
                        levels = c("1", "0.95", "0.75", "0.5"),
                        labels = c("100%", "95%", "75%", "50%"))) %>% 
  ggplot() +
  geom_sf(aes(color = level), 
          fill = NA, linewidth = 2) +
  geom_point(data = mcps$data, aes(x = x_, y = y_),
             color = "gray30", alpha = 0.5) +
  xlab(NULL) +
  ylab(NULL) +
  scale_color_brewer(name = "MCP Level",
                     palette = "Set1") +
  theme_bw() +
  theme(legend.position = "bottom", 
        legend.box.background = element_rect(colour = "black", linewidth = 0.8))

hr_overlap(mcps,kdes)
```


c. Autocorrelated Kernel Density estimator of home range

```{r}

# Without going into detail on the different CTMMs, we'll demonstrate fitting
# an aKDE with an Ornstein-Uhlenbeck model.
akdes <- hr_akde(trk_108, model = fit_ctmm(trk_108, "ou"), levels = 0.95)

# Get area in km^2
hr_area(akdes, units = TRUE) %>% 
  mutate(area = units::set_units(area, "km^2"))

# plot the polygon isopleths by themeself
plot(hr_isopleths(akdes)[1])

# Plot the isopleth with data on top
plot(akdes) 

# Custom plot with ggplot2 (thanks to Brian Smith for this code chunk)

hr_isopleths(akdes) %>% 
  # Make level a factor for discrete color scales
  # Can control order and labels here
  mutate(level = factor(level, 
                        levels = c("1", "0.95", "0.75", "0.5"),
                        labels = c("100%", "95%", "75%", "50%"))) %>% 
  ggplot() +
  geom_sf(aes(color = level), 
          fill = NA, linewidth = 2) +
  geom_point(data = mcps$data, aes(x = x_, y = y_),
             color = "gray30", alpha = 0.5) +
  xlab(NULL) +
  ylab(NULL) +
  scale_color_brewer(name = "MCP Level",
                     palette = "Set1") +
  theme_bw() +
  theme(legend.position = "bottom", 
        legend.box.background = element_rect(colour = "black", linewidth = 0.8))

hr_overlap(kdes, akdes)

```


3. Explore the movement of all animals in the data set extracting a continuous covariate
(for example elevation or distance to roads) to the data. Extract the covariate both at the end points and along the steps. Explore the differences between the two ways of extracting the data, for example fitting a linear regression with step length as response variable and the extracted variable as explanatory variable.

```{r}
library(terra)

airport <- terra::rast("../data/US_VFR_Wall_Planning_Chart.tif")
airport

names(airport) <- "airport"
airport

plot(airport, main = "airport")

airport_colors <- coltab(airport)[[1]]

# blues
airport_colors[airport_colors$red == 20,]

# yellows
airport_colors[airport_colors$red >= 249,]

# reds
airport_colors[airport_colors$red == 136,]

airport_colors$isAirport <- ifelse(airport_colors$value %in% c(1,8,10,28,68), TRUE, FALSE)
airport_class <- cbind(airport_colors$value, airport_colors$isAirport)
colnames(airport_class) <- c("code","isAirport")
airport_class <- as.tibble(airport_class)


```


4. Fit a habitat selection function of your own choice (resource selection function at one scale or an (integrated) step-selection function) to the data using one covariate.

